Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-11-26 patchy-review-responses.tex

\begin{quote}
As can you see from the reviews, two general points were raised by all
reviewers and I feel that thoroughly addressing both of these points poses a
prerequisite for further consideration of the manuscript:

(i) There needs to be some sort of numerical verification (e.g. with
individual-based simulations) to test the main conclusions, check their
robustness, and identify the parameter ranges for which they hold.

(ii) The presentation of the main result needs to be improved. In particular,
all reviewers feel that the main result can be derived in a simpler and more
intuitive way. I agree with them that at present the manuscript is overly
technical and would benefit strongly from a more intuitive derivation of the
results and discussion of the assumptions under which they hold. The reviewers
provide detailed suggestions on how this can be achieved.
\end{quote}

Changes of note are:

\begin{itemize}

  \item We have done a large number of simulations, and compared the results to the theory;
    see figures~\ref{fig:sim_probs} and \ref{fig:sim_times}.

  \item Several reviewers pointed out that in approximating $\E[1-(1-p_e)^K] \approx \E[K] p_e$
    we were neglecting the case that $s_m \ll s_p$.
    This was pur laziness on our part,
    which we have rectified; see the new equation~\eqref{eqn:twocase_migrate}.

\end{itemize}

@@

%%%%%%
\point{}{
    Am I right that the basic argument is the
    following:

    A migrant family in unfavorable territory will be typically of size $n \sim 1/s$
    ($s=s_m$), stick around for a time $\tau \sim 1/s$ and be spread out over $\ell \sim \sigma/\sqrt(s)$. 
    Given an equilibrium density $\rho*q(x)$, we need an arrival rate
    of migrant families of
    \[ k ~ \rho q(x) l/n/\tau \sim \rho q(x) l s^2 \]
}

\reply
The factor of $\ell$ shouldn't be there:
suppose that $X$ is a Poisson process of ``centers'' of rate 1, 
and that around each center of $X$ we distribute $n$ points uniformly within a circle of radius $\ell$;
the expected number of points in a set of area 1 is $n$, regardless of $\ell$.
The right thing to do with the factor of $\tau$ is harder to intuit:
it is clear that $\rho q(x)$ should be $E[K]$ multiplied by the number of families present;
and it might be argued that the number of families present should be the influx of families
multiplied by the time they stick around for,
but given that the typical size of a family decreases once it arrives,
the correct way forward becomes non-intuitive, to us, at least.

%%%%%%%%
\point{}{ 
    The probability of establishment of at least one member of family is then
    \[ P_{est} = 1 - \exp(-p n) \sim 1 - \exp(-p_e/s) \]
    where $p_e$ is the establishment probability of a single individual. The result
    for $\migrate$ will then depend on whether $p_e/s$ is large or small.
    For $p_e/s$ large, $\migrate = \rho q(x) s^2 \ell$,
    otherwise,
    $\migrate = \rho q(x) s \ell p_e$.

@@

    The latter is the case discussed in the manuscript. The other is important too
    and should be discussed. 
}

\reply
As discussed above, we have incorporated this good point; see around equation~\eqref{eqn:twocase_migrate}.

@@

%%%%%%%%
\point{}{ 
    To assess the accuracy of the various approximations made, the process should
    be simulated and the measured rates explicitly compared to analytical
    predictions.
}

\reply
  We've done this; see above.

%%%%%%%%
\point{long range migration}{All results depend exponentially on the distance, which
      is a consequence of the model only allowing for short range dispersal. Rare
    long range dispersal will alter all these rates dramatically, see for example
    recent work by Hallatschek and Fisher.
}

\reply
This was already discussed in the discussion, but we've added a note about it earlier on, citing Hallatschek \& Fisher.
\llname{rr:longtailed}

@@

\reply

%%%%%%%%%%%%%%  Reviewer 3
\reviewersection

\point{1a.}{
        What I miss is a clearer discussion of the scope and limits of the
    main results. Take Eq. 13: in its revised form, this is a non-monotonic
    function of the selection coefficient $s_m$. I am almost sure that this is
    an artefact and that the parameter range of the theory needs to be
    restricted to the part where the transition probability declines with $s_m$.
    But it is not easy to see how this needs to be done in a proper way. We
    certainly need a minimal distance among patches. On top of that, we may
    need a lower bound for $s_m$. Both these restrictions (or any other) should
    not be just ad-hoc, but motivated from within the theory (see my remarks at eq. 13).
}

\reply
Very good points.  
We apologize for not catching the nonmonotonicity earlier.
We've added a section, \nameref{ss:assumptions}, discussing this point,
as well as a note regarding the nonmonotonicity wrt $s_m$
\llname{rr:sm_condition}.

### 2014-11-26 patchy-selection-paper.tex
more sims

Also, to avoid introducing more parameters,
we assume population density $\rho$ is equal across the range (even in the ``poor'' patches).
If this is not the case, in most equations below, the population density that should be used is the density
on the poor patches; see \citet{lenormand2002limits} for a more thorough approach.
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance). 
We will deal with migration by immediately appealing to the central limit theorem,
treating the total distance traveled across $t$  dispersal events as Gaussian with variance $t \sigma^2$,
and do not discuss the (interesting) cases where rare, long-distance dispersal events
are more important \citep[see][for discussion]{longtailed,ralphcoop2010,hallatschek??}.  \linelabel{rr:longtailed}
%We also aim to describe the scenario where it is sensible to think of large, discrete patches;
%the case of a fine-scale heterogeneous environment is treated
%\citep{elsewhere}. \gc{IS THIS LAST SENTENCE NEEDED IF SO GIVE CITATION} \plr{haven't found it yet}

@@

This suggests that 
\begin{align} \label{eqn:ratio_K}
  \migrate(x) 
    &= \rho q(x) \times 
    \frac{
        \E[1-(1-p_e)^K]
    }{
        \E[K]/2s_m
    } .
\end{align}
There are now two extremes to consider.
If $\P\{K>1/p_e\}$ is small, then each family has a small chance of establishment,
and the numerator is approximately $p_e \E[K]$.
On the other hand, if $p_e$ is not too small,
then the numerator would be close to 1, as any family that arrives will likely establish.
Combining these, and absorbing the term $\E[W|W>1]$ from equation \eqref{eqn:mean_K} 
into the constant $C$,
\begin{align} \label{eqn:twocase_migrate}
    \migrate(x) &\approx C \rho s_m \min(s_m,p_e) \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma) .
%     \begin{cases}
%         q(x) \times 4 \rho s_m s_p / \xi^2  \qquad & \text{ if } s_p < s_m \\
%         q(x) \times 2 \rho s_m^2 \qquad & \text{ if } s_p > s_m 
%     \end{cases} .
\end{align}
When we use this, we set $p_e = 2 s_p / \xi^2$.

@@

One way to visualize this is by a phase diagram.
In figure \ref{fig:phase_diagram},
we give a contour plot for the time until adaptation by migration ($\Tmig = 1/\migrate$) and by new mutation ($\Tmut=1/\mutrate$)
as a function of various parameters. 
This gives us an idea of not only the parameter space where parallel adaptation is likely ($\Tmig \ll \Tmut$),
but also where adaptation by any means takes a very long time ($\Tmig$ and $\Tmut$ both very large),
an important check in applications of these results.

% \begin{figure}[ht]
%   \begin{center}
%     \includegraphics{phase-diagram-log}
%   \end{center}
%   \caption{
%   Shaded contour plots of
%   mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
%   as a function of population density ($\rho$), 
%   distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
%   strength of deleterious selection ($s_m$),
%   and mutation rate scaled by patch area ($\mu A$).
%   The right-hand plot shows that $\migrate$ is relatively insensitive to $s_m$ (thanks to the square root).
%   The black, dotted line shows values where $\Tmig=\Tmut$.
%   Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=20$, and $\xi^2=1$.
%   Note that population density $\rho$ and mutation influx $A\mu$ are shown on a log scale.
%   \label{fig:phase_diagram}
%   }
% \end{figure}

@@

\paragraph{Long distance migration}
We have also ignored the possibility of very long distance migration,
instead focusing on local dispersal (and using Gaussians via the central limit theorem).
However, dispersal distributions can be very heavy tailed, 
with a small fraction of individuals moving very long distances indeed \citep{levin2003ecology,reynolds2009levy}.
In addition, over long time-scales, very rare chance events (mice carried off by hurricanes and the like \citet{censky1998overwater,nathan2008mechanisms})
could play a role in spreading migrant alleles if adaptation by other means is sufficiently unlikely.
Such tail events could greatly increase the probability of parallel adaptation above that predicted by our model. 
Furthermore, if adaptive alleles do move between distant patches via rare, long distance migration 
then they will be associated with a much longer shared haplotype than predicted by equation \eqref{eqn:haplotype_length}. 
As such, we view our results as a null model by which the contribution of long distribution migrants to adaptation could be empirically judged.

>>>>

Also, to avoid introducing more parameters,
we assume population density $\rho$ is equal across the range (even in the ``poor'' patches).
If this is not the case, in most equations below, the population density that should be used is the density
on the poor patches; see \citet{lenormand2002limits} for a more thorough approach.
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance). 
We will deal with migration by immediately appealing to the central limit theorem,
treating the total distance traveled across $t$  dispersal events as Gaussian with variance $t \sigma^2$,
and do not discuss the (interesting) cases where rare, long-distance dispersal events
are more important \citep[see e.g.][for discussion]{levin2003ecology,ralph2010parallel,hallatschek2014acceleration}.  \linelabel{rr:longtailed}
%We also aim to describe the scenario where it is sensible to think of large, discrete patches;
%the case of a fine-scale heterogeneous environment is treated
%\citep{elsewhere}. \gc{IS THIS LAST SENTENCE NEEDED IF SO GIVE CITATION} \plr{haven't found it yet}

@@

This suggests that 
\begin{align} \label{eqn:ratio_K}
  \migrate(x) 
    &= \rho q(x) \times 
    \frac{
        \E[1-(1-p_e)^K]
    }{
        \E[K]/2s_m
    } .
\end{align}
There are now two extremes to consider.
If $\P\{K>1/p_e\}$ is small, then each family has a small chance of establishment,
and the numerator is approximately $p_e \E[K]$.
On the other hand, if $p_e$ is not too small,
then the numerator would be close to 1, as any family that arrives will likely establish.
Combining these, and absorbing the term $\E[W|W>1]$ from equation \eqref{eqn:mean_K} 
into the constant $C$,
\begin{align} \label{eqn:twocase_migrate}
  \migrate(x) &\approx C \rho s_m \min(s_m,p_e) \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma)  \qquad \text{for } |y| > \sigma/\sqrt{2 s_m} 
%     \begin{cases}
%         q(x) \times 4 \rho s_m s_p / \xi^2  \qquad & \text{ if } s_p < s_m \\
%         q(x) \times 2 \rho s_m^2 \qquad & \text{ if } s_p > s_m 
%     \end{cases} .
\end{align}
When we use this below, we set $p_e = 2 s_p / \xi^2$.
Note that the function increases for small $s_m$, 
but decreases if $s_m > |y|^2 / 2 \sigma^2$, 
i.e.\ where the condition holds.

@@

% One way to visualize this is by a phase diagram.
% In figure \ref{fig:phase_diagram},
% we give a contour plot for the time until adaptation by migration ($\Tmig = 1/\migrate$) and by new mutation ($\Tmut=1/\mutrate$)
% as a function of various parameters. 
% This gives us an idea of not only the parameter space where parallel adaptation is likely ($\Tmig \ll \Tmut$),
% but also where adaptation by any means takes a very long time ($\Tmig$ and $\Tmut$ both very large),
% an important check in applications of these results.
% 
% \begin{figure}[ht]
%   \begin{center}
%     \includegraphics{phase-diagram-log}
%   \end{center}
%   \caption{
%   Shaded contour plots of
%   mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
%   as a function of population density ($\rho$), 
%   distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
%   strength of deleterious selection ($s_m$),
%   and mutation rate scaled by patch area ($\mu A$).
%   The right-hand plot shows that $\migrate$ is relatively insensitive to $s_m$ (thanks to the square root).
%   The black, dotted line shows values where $\Tmig=\Tmut$.
%   Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=20$, and $\xi^2=1$.
%   Note that population density $\rho$ and mutation influx $A\mu$ are shown on a log scale.
%   \label{fig:phase_diagram}
%   }
% \end{figure}

@@

\paragraph{Long distance migration}
We have also ignored the possibility of very long distance migration,
instead focusing on local dispersal (and using Gaussians via the central limit theorem).
However, dispersal distributions can be very heavy tailed, 
with a small fraction of individuals moving very long distances indeed \citep{levin2003ecology,reynolds2009levy}.
In addition, over long time-scales, very rare chance events (mice carried off by hurricanes and the like; \citet{censky1998overwater,nathan2008mechanisms})
could play a role in spreading migrant alleles if adaptation by other means is sufficiently unlikely.
Such tail events could greatly increase the probability of parallel adaptation above that predicted by our model. 
Furthermore, if adaptive alleles do move between distant patches via rare, long distance migration 
then they will be associated with a much longer shared haplotype than predicted by equation \eqref{eqn:haplotype_length}. 
As such, we view our results as a null model by which the contribution of long distribution migrants to adaptation could be empirically judged.
